cardinalR: Generating interesting high-dimensional data structures

### 2025-04-20 paper-cardinalR.Rmd
Added intro

However, a research gap exists in the comprehensive and flexible generation of high-dimensional data structures that seamlessly integrate fundamental geometric principles with sophisticated noise control and customizable cluster generation. While `geozoo` offers geometric primitives, it lacks direct mechanisms for extending these structures to arbitrary high dimensions with controlled noise or for generating clustered data based on these shapes with user-defined parameters like position, scale, and orientation. `snedata` focuses on creating abstract datasets for dimensionality reduction evaluation, not on building high-dimensional data directly from geometric layouts with flexible noise and clustering options. `mlbench` provides static benchmark datasets, limiting the ability to generate data with specific, user-defined geometric structures and noise profiles in high dimensions. Similarly, `splatter` is specialized for biological data simulation and does not offer the general framework for creating diverse geometric structures in arbitrary dimensions with fine-grained control over noise and cluster arrangements based on basic geometric shapes.

To address this gap, this paper introduces the `cardinalR` R package. This novel package provides a versatile suite of functions designed to generate customizable data structures in any number of dimensions, starting from fundamental geometric primitives. `cardinalR` offers key functionalities that go beyond the capabilities of existing tools by enabling users to: (i) construct high-dimensional datasets based on geometric shapes, with the ability to augment dimensionality through the addition of controlled noise dimensions; (ii) introduce tunable levels of background noise to these structures; and (iii) generate complex clustered data arrangements by leveraging basic geometric forms with precise control over their positions, scales, orientations, and sample sizes in arbitrary dimensional spaces. By providing these integrated functionalities, `cardinalR` aims to empower researchers to create more nuanced and challenging synthetic datasets tailored to the specific demands of evaluating algorithms in high-dimensional settings and complex spatial scenarios, bridging the gap between geometric foundations and the flexible generation of complex synthetic data.

>>>>

There is a valuable opportunity to improve the generation of high-dimensional data structures by integrating geometric principles with advanced noise control and customizable clustering. The `geozoo` package provides a strong foundation but could be enhanced to support high-dimensional extensions with controlled noise and user-defined parameters for clustering. Similarly, while `snedata` focuses on abstract datasets for dimensionality reduction, adding features for generating high-dimensional data from geometric layouts would enhance its usability. The `mlbench` package could also benefit from allowing users to create datasets with specific geometric structures and noise profiles. Additionally, although `splatter` specializes in biological data simulation, it could be expanded to offer a broader framework for generating diverse geometric structures across dimensions, enabling detailed control over noise and clustering. Addressing these areas could lead to more robust high-dimensional data generation tools.

To address this gap, this paper introduces the `cardinalR` R package. This package provides a collection of functions designed to generate customizable data structures in any number of dimensions, starting from basic geometric shapes. `cardinalR` offers important functionalities that extend beyond the capabilities of existing tools, allowing users to: (i) construct high-dimensional datasets based on geometric shapes, including the option to enhance dimensionality by adding controlled noise dimensions; (ii) introduce adjustable levels of background noise to these structures; and (iii) create complex clustered data arrangements by using fundamental geometric forms, while maintaining their positions, scales, orientations, and sample sizes in arbitrary dimensional spaces. By providing these integrated features, `cardinalR` aims to provide researchers to generate more explainable and challenging synthetic datasets focused to the specific needs of evaluating algorithms in high-dimensions. This bridges the gap between geometric foundations and the flexible generation of complex synthetic data.
